---
title: "Smart Device Fitness Data Trends"
subtitle: "Unlocking growth opportunities for Bellabeat"
author: "Presentation by: Emeka Phil-ebosie"
date: "1/22/2022"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```



```{r Library Loading, results='hide', message=FALSE, echo = FALSE}
# Loading the necessary packages. 

library("tidyverse")
library("lubridate")
library("here")
library("janitor")
library("dplyr")
library("ggplot2")
library("RColorBrewer")
library("plotly")
```


```{r Data loading and cleaning, message=FALSE, echo=FALSE, results='hide'}
# This section loads the data and does some  cleaning before processing. 
# Preliminary inspections are done to determine the metrics logged. 
# Sum or average (which is applicable) is created for each user for each day in the overview_df
# Averages per day are then created with in the daily_average_df
 
df <- 
  read.csv("../Fitabase Data 4.12.16-5.12.16/dailyActivity_merged.csv")
df$ActivityDate <- as.Date(df$ActivityDate, "%m/%d/%Y")
distance_df <- df
daily_user_df <-
  distance_df %>% 
  drop_na(TotalDistance) %>%
  group_by(user_id = Id, activity_date = floor_date(ActivityDate, "day")) %>%
  summarize(total_distance = sum(TotalDistance))
overview_df <- daily_user_df
daily_average_df <-
  daily_user_df %>%
  group_by(activity_date) %>%
  summarize(av_distance = mean(total_distance))

df <- 
  read.csv("../Fitabase Data 4.12.16-5.12.16/minuteCaloriesNarrow_merged.csv")
df$ActivityMinute <- as.Date(df$ActivityMinute, "%m/%d/%Y")
calories_df <- df %>%
  rename(ActivityDate = ActivityMinute)
daily_user_df  <-
  calories_df %>% 
  drop_na(Calories) %>%
  group_by(user_id = Id, activity_date = floor_date(ActivityDate, "day")) %>%
  summarize(total_calories = sum(Calories))
overview_df <- 
  left_join(overview_df, daily_user_df, by=c("user_id","activity_date"))
df_for_merging <-
  daily_user_df %>%
  group_by(activity_date) %>%
  summarize(av_calories = mean(total_calories))
daily_average_df <- 
  left_join(daily_average_df, df_for_merging, by=c("activity_date"))
  

df <- 
  read.csv("../Fitabase Data 4.12.16-5.12.16/minuteIntensitiesNarrow_merged.csv")
df$ActivityMinute <- as.Date(df$ActivityMinute, "%m/%d/%Y")
intensities_df <- df %>%
  rename(ActivityDate = ActivityMinute)
daily_user_df <-
  intensities_df %>%  
  drop_na(Intensity) %>%
  group_by(user_id = Id, activity_date = floor_date(ActivityDate, "day")) %>%
  summarize(total_intensity = sum(Intensity))
overview_df <- 
  left_join(overview_df, daily_user_df, by=c("user_id","activity_date"))
df_for_merging <-
  daily_user_df %>%
  group_by(activity_date) %>%
  summarize(av_intensity = mean(total_intensity))
daily_average_df <- 
  left_join(daily_average_df, df_for_merging, by=c("activity_date"))

df <- 
  read.csv("../Fitabase Data 4.12.16-5.12.16/minuteStepsNarrow_merged.csv")
df$ActivityMinute <- as.Date(df$ActivityMinute, "%m/%d/%Y")
steps_df <- df %>%
  rename(ActivityDate = ActivityMinute)
daily_user_df <-
  steps_df %>%  
  drop_na(Steps) %>%
  group_by(user_id =Id, activity_date = floor_date(ActivityDate, "day")) %>%
  summarize(total_steps = sum(Steps))
overview_df <- 
  left_join(overview_df, daily_user_df, by=c("user_id","activity_date"))
df_for_merging <-
  daily_user_df %>%
  group_by(activity_date) %>%
  summarize(av_steps = mean(total_steps))
daily_average_df <- 
  left_join(daily_average_df, df_for_merging, by=c("activity_date"))

df <- 
  read.csv("../Fitabase Data 4.12.16-5.12.16/minuteMETsNarrow_merged.csv")
df$ActivityMinute <- as.Date(df$ActivityMinute, "%m/%d/%Y")
mets_df <- df %>%
  rename(ActivityDate = ActivityMinute)
daily_user_df <-
  mets_df %>%  
  drop_na(METs) %>%
  group_by(user_id = Id, activity_date = floor_date(ActivityDate, "day")) %>%
  summarize(av_mets = mean(METs))
overview_df <- 
  left_join(overview_df, daily_user_df, by=c("user_id","activity_date"))
df_for_merging <-
  daily_user_df %>%
  group_by(activity_date) %>%
  summarize(av_mets = mean(av_mets))
daily_average_df <- 
  left_join(daily_average_df, df_for_merging, by=c("activity_date"))

df <- 
  read.csv("../Fitabase Data 4.12.16-5.12.16/sleepDay_merged.csv")
df$SleepDay <- as.Date(df$SleepDay, "%m/%d/%Y")
sleep_df <- df %>%
  rename(ActivityDate = SleepDay)
daily_user_df <-
  sleep_df %>%  
  drop_na(TotalMinutesAsleep) %>%
  group_by(user_id = Id, activity_date = floor_date(ActivityDate, "day")) %>%
  summarize(total_sleep = sum(TotalMinutesAsleep)/60)
overview_df <- 
  left_join(overview_df, daily_user_df, by=c("user_id","activity_date"))
df_for_merging <-
  daily_user_df %>%
  group_by(activity_date) %>%
  summarize(av_sleep = mean(total_sleep))
daily_average_df <- 
  left_join(daily_average_df, df_for_merging, by=c("activity_date"))


df <- 
  read.csv("../Fitabase Data 4.12.16-5.12.16/heartrate_seconds_merged.csv")
df$Time <- as.Date(df$Time, "%m/%d/%Y")
heartrate_df <- df %>%
  rename(ActivityDate = Time)
daily_user_df <-
  heartrate_df %>%  
  drop_na(Value) %>%
  group_by(user_id = Id, activity_date = floor_date(ActivityDate, "day")) %>%
  summarize(av_heartrate = mean(Value))
overview_df <- 
  left_join(overview_df, daily_user_df, by=c("user_id","activity_date"))
df_for_merging <-
  daily_user_df %>%
  group_by(activity_date) %>%
  summarize(av_heartrate = mean(av_heartrate))
daily_average_df <- 
  left_join(daily_average_df, df_for_merging, by=c("activity_date"))

df <- 
  read.csv("../Fitabase Data 4.12.16-5.12.16/weightLogInfo_merged.csv")
df$Date <- as.Date(df$Date, "%m/%d/%Y")
weight_df <- df %>%
  rename(ActivityDate = Date)
weight_df <- weight_df[ -c(4,5,7,8)]
daily_user_df <-
  weight_df %>%  
  drop_na(BMI) %>%
  group_by(user_id = Id, activity_date = floor_date(ActivityDate, "day")) %>%
  summarize(av_weight = mean(WeightKg), av_BMI = mean(BMI))
overview_df <- 
  left_join(overview_df, daily_user_df, by=c("user_id","activity_date"))
df_for_merging <-
  daily_user_df %>%
  group_by(activity_date) %>%
  summarize(av_weight = mean(av_weight), av_BMI = mean(av_BMI))
daily_average_df <- 
  left_join(daily_average_df, df_for_merging, by=c("activity_date"))


```


## Agenda

* Business Task Summary
* Available Smart Device Data
* Activity Level Trend
* Findings
* Activity Level Metrics
* Recommendation
* Appendix 


## Business Task Summary
#### Identifying growth opportunities from trends

__Bellabeat Company__

* A high-tech manufacturer of health-focused products for women.
* Products provide users information on activity, sleep, stress, menstrual cycle, and mindfulness habits. 

__Business Task Summary__

* This analysis should identify user trends from publicly available fitness smart device data to guide Bellabeat’s marketing strategy and unlock growth opportunities. 


## Available Smart Device Data 
#### Public Data from FitBit users was evaluated 

* Public FitBit data not obtained directly by Bellabeat was used. 
* 33 eligible Fitbit users consented to the use of 1 month of personal tracker information.
* Data tracked included distance, steps, intensity, calories and METs.

```{r Activity Charts, echo=FALSE, message=FALSE, warning=FALSE}

# Introductory chart to show the overview 
# Initially wanted to start with a pie chart. 
# However horizontal bar chart is an easier introductory chart IMO. 
# piechart_df gets the column names and their number of values from the overview_df.
# Additional formatting is done to use the appropriate names.

piechart_df <- data.frame(t(colSums(!is.na(overview_df)))) %>%
  select(-user_id, -activity_date) %>%
  pivot_longer(cols = names(.))  %>% 
  separate(name, into = c("alfa", "name"), sep ="_") %>%
  select(name, value)

piechart_df$name[1:8] <- str_to_title(piechart_df$name[1:8])
piechart_df$name[5] <- "METs"

piechart_df <- piechart_df %>% 
  mutate(prop = value / sum(piechart_df$value) *100) %>%
  mutate(prop_round = round(prop)) 

# Commented out the pie chart as mentioned above. May be useful later. 

# plot_ly(type='pie', labels=piechart_df$name , 
#        values= piechart_df$prop_round, textinfo='label+percent',
#        showlegend = FALSE, 
#        textfont = list(size = 16),
#        title = 'Average Usage of Fitbit Smart Device',
#        titlefont = list(size =25))


color = brewer.pal(length(piechart_df$name),"Set1")
ggplot(piechart_df, aes(x = value, y = reorder(name, value))) +
  geom_col(fill = color)+
  theme(panel.background = element_rect(fill="white"),
        axis.ticks = element_blank(),
        axis.text = element_text(size=10),
        axis.title = element_text(size=12, face="bold"))+
  xlab("Total user records") + ylab("Metrics")+
  labs(title = "Metrics Tracked by Users",
       subtitle = "Total number of main metrics tracked daily for one month",
       caption = "Public Domain, dataset made available through Mobius on Kaggle")


```


## Activity Level Trend
### High threshold of activity maintained

* Users consistently maintained a high level of activity for the period. 
* For example, analysis revealed an average of above 7,500 steps per day. 
* Similar performance levels are observed in other metrics e.g. METs, intensity, and sleep.

```{r Step Chart with Trend, message=FALSE, warning=FALSE}

# Used ggplot to for the steps chart. Added some annotations to emphasize the trend. 
# Outlier is excluded because the last the day data is spurious for most metrics. 
# Appendix shows the outlier. 



daily_average_df %>%
  filter(activity_date != max(activity_date)) %>%
  ggplot(aes(x=activity_date, y=av_steps))+
  geom_point(color="blue")+
  ylim(0,9000)+
  stat_smooth(method = "lm", se = FALSE)+
  theme(panel.background = element_rect(fill="white", colour = "grey50"),
        axis.text = element_text(size=10),
        axis.title = element_text(size=12, face="bold"))+
  xlab("Day") + ylab("Average number of steps")+
  labs(title = " Daily Average Steps",
       subtitle = "Mean number of steps logged by users daily",
       caption = "Public Domain dataset made available through Mobius on Kaggle")+ 
  annotate("segment", x = as.Date(c("2016-04-25")), xend = as.Date(c("2016-04-20")), 
           y = 5000 , yend = 7400,
           colour = "red", size=2, alpha=0.5, arrow=arrow())+
  annotate("text", size=5, x = as.Date(c("2016-04-25"))+2, y = 5000-400, 
           label = "Average performance trends above 7500 steps per day")

```


## Findings 
### Calories strongly correlates with increased activity

* Calories burned make up a large amount (~20%) of metrics logged by users.
* And higher steps corresponds to higher calories burned. 
 
```{r Calories,  message=FALSE, warning=FALSE}

# Used plot_ly to generate a simple pie chart from the piechart_df
# Used ggplot with annotations to show positive trend. 
# Outlier is excluded because the last the day data is spurious for most metrics. 
# Appendix shows the outlier. 

name <- c(piechart_df$name[2], "Others")
prop <- c(piechart_df$prop_round[2], 100-piechart_df$prop_round[2])
df <- data.frame(name, prop)


plot_ly(type='pie', labels=df$name, 
        values= df$prop, textinfo='label+percent',
        showlegend = FALSE, 
        textfont = list(size = 25),
        title = 'Calories Logged Compared to Other Metrics',
        titlefont = list(size =25))

daily_average_df %>%
  filter(activity_date != max(activity_date)) %>%
  ggplot(aes(x=av_steps, y=av_calories))+
  geom_point(color="blue")+
  stat_smooth(method = "lm", se = FALSE)+
  theme(panel.background = element_rect(fill="white", colour = "grey50"),
        axis.text = element_text(size=10),
        axis.title = element_text(size=12, face="bold"))+
  xlab("Steps") + ylab("Calories")+
  labs(title = " Calories Per Steps",
       subtitle = "Increase in steps with higher calories",
       caption = "Public Domain, dataset made available through Mobius on Kaggle")+
  annotate("segment", x = 6600, xend = 7800, 
           y = 2330 , yend = 2430,
           colour = "red", size=2, alpha=0.5, arrow=arrow())


```


## Activity Level Metrics
### Strong positive correlation confirmed with calories

* The positive relationship between calories burned and higher activity level is evident. 
* This is confirmed with the other metrics such as distance, intensity and METs.  
 
```{r Plotting the parameters, message=FALSE, warning=FALSE, fig.show="hold", out.width="50%", fig.cap="Public Domain dataset made available through Mobius on Kaggle. " }

# Used ggplot to show matching trend of steps with calories, METs, and Intensity. 
# Outlier is excluded because the last the day data is spurious for most metrics. 
# Appendix shows the outlier. 

daily_average_df %>%
  filter(activity_date != max(activity_date)) %>%
  ggplot(aes(x=av_distance, y=av_calories))+
  geom_point(color="blue")+
  stat_smooth(method = "lm", se = FALSE)+
  theme(panel.background = element_rect(fill="white", colour = "grey50"),
        axis.text = element_text(size=10),
        axis.title = element_text(size=12, face="bold"))+
  xlab("Distance") + ylab("Calories")+
  labs(title = "Distance vs. Calories")

daily_average_df %>%
  filter(activity_date != max(activity_date)) %>%
  ggplot(aes(x=av_intensity, y=av_calories))+
  geom_point(color="blue")+
  stat_smooth(method = "lm", se = FALSE)+
  theme(panel.background = element_rect(fill="white", colour = "grey50"),
        axis.text = element_text(size=10),
        axis.title = element_text(size=12, face="bold"))+
  xlab("Intensity") + ylab("Calories")+
  labs(title = "Intensity vs. Calories")

daily_average_df %>%
  filter(activity_date != max(activity_date)) %>%
  ggplot(aes(x=av_mets, y=av_calories))+
  geom_point(color="blue")+
  stat_smooth(method = "lm", se = FALSE)+
  theme(panel.background = element_rect(fill="white", colour = "grey50"),
        axis.text = element_text(size=10),
        axis.title = element_text(size=12, face="bold"))+
  xlab("METs") + ylab("Calories")+
  labs(title = "METs vs. Calories")


```


## Recommendation
### Way forward

* Bellabeat’s marketing strategy should highlight the strong relationship between calories burned and high activity levels. 
* Focus should also emphasize how Bellabeat product line can help track calories burned and activity levels with discreet fashionable products like Leaf and Time. 
* Further analysis should be carried out with Bellabeat’s smart device fitness data over a longer period of time to confirm this and other trends with its target demographic. 


## The End


## Appendix
### Outlier on the last day excluded from analysis

```{r Showing outliers, message=FALSE, warning=FALSE, fig.show="hold", out.width="50%" }


# Plotting identified parameters to identify outliers. 

ggplot(data=daily_average_df)+
  geom_point(mapping = aes(x=activity_date, y=av_calories), color="blue")+
  theme(panel.background = element_rect(fill="white", colour = "grey50"),
        axis.text = element_text(size=10),
        axis.title = element_text(size=12, face="bold"))+
  ggtitle("Daily Average Calories")+
  xlab("Day") + ylab("Average calories per day")+
  annotate("text", x = max(daily_average_df$activity_date)-2, y = 1220, label = "Outlier")

ggplot(data=daily_average_df)+
  geom_point(mapping = aes(x=activity_date, y=av_intensity), color="blue")+
  theme(panel.background = element_rect(fill="white", colour = "grey50"),
        axis.text = element_text(size=10),
        axis.title = element_text(size=12, face="bold"))+
  ggtitle("Daily Average Intensity")+
  xlab("Day") + ylab("Average intensity per day")+
  annotate("text", x = max(daily_average_df$activity_date)-2, y = 125, label = "Outlier")

ggplot(data=daily_average_df)+
  geom_point(mapping = aes(x=activity_date, y=av_steps), color="blue")+
  theme(panel.background = element_rect(fill="white", colour = "grey50"),
        axis.text = element_text(size=10),
        axis.title = element_text(size=12, face="bold"))+
  ggtitle("Daily Average Steps")+
  xlab("Day") + ylab("Average number of steps")+
  annotate("text", x = max(daily_average_df$activity_date)-2, y = 3600, label = "Outlier")


ggplot(data=daily_average_df)+
  geom_point(mapping = aes(x=activity_date, y=av_mets), color="blue")+
  theme(panel.background = element_rect(fill="white", colour = "grey50"),
        axis.text = element_text(size=10),
        axis.title = element_text(size=12, face="bold"))+
  ggtitle("Daily Average METs")+
  xlab("Day") + ylab("Average METs (Active minutes)")+
  annotate("text", x = max(daily_average_df$activity_date)-2, y = 13.35, label = "Outlier")



```


