---
title: "NCDC Nigeria Covid Data Web Scrape"
author: "pebosie"
date: "1/29/2022"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```



```{r Library Loading, results='hide', message=FALSE, echo = FALSE}
# Loading the necessary packages. 

library("rvest")
library("tidyverse")
library("readr")
library("tidyverse")
library("lubridate")
library("janitor")
library("dplyr")
library("ggplot2")
library("RColorBrewer")
library("plotly")
library("naijR")
```



```{r Data loading and cleaning, message=FALSE, echo=FALSE, results='hide'}
# This section scrapes the data table and additional text from the website. 
# Cleans the data to make it ROCCC.  
# Text data is also parsed into a usable data frame

# Reading the table. 
url_data <- read_html("https://covid19.ncdc.gov.ng/")
css_selector <- "#custom1"
states_df <- url_data %>%
  html_element(css=css_selector) %>%
  html_table()
states_df <- states_df %>%
  rename(states_affected = `States Affected`,
        lab_confirmed_cases = `No. of Cases (Lab Confirmed)`,
        admission_cases =  `No. of Cases (on admission)`,
        discharged =  `No. Discharged`,
        deaths = `No. of Deaths`)
states_df$lab_confirmed_cases <- as.integer(gsub(",","",states_df$lab_confirmed_cases)) 
states_df$admission_cases <- as.integer(gsub(",","",states_df$admission_cases))
states_df$discharged <- as.integer(gsub(",","",states_df$discharged)) 
states_df$deaths <- as.integer(gsub(",","",states_df$deaths)) 



# Reading additional text upate
highlights_df <- url_data %>%
  html_nodes(".temp") %>%
  html_text()


df <- url_data %>% 
  html_nodes(".card-box")  %>% 
  html_text()
hl_df <- as.data.frame(str_split(df[1], "\n"))
colnames(hl_df) <- c("Highlights")
hl_df$Highlights <- trimws(hl_df$Highlights, which = c("both"))
hl_arr <- hl_df[!apply(hl_df == "", 1, all), ]
mon_arr <- array(month.name)
month_match <- str_c(mon_arr, collapse = "|")
date_sentence <- str_subset(hl_arr[1:3], month_match)
matches <- str_extract(date_sentence, month_match)
y_loc <- str_locate(date_sentence,matches)
pub_date <- substr(date_sentence, start = y_loc[1,1]-8, stop = y_loc[1,2]+5)
pub_date <- trimws(pub_date, which =c("both"))


# Get the states that affected for that day.
state_match <- states_df$states_affected
state_match <- str_c(state_match, collapse = "|")
news_sentence <- str_subset(hl_arr[3:5], state_match)
States <- str_extract_all(news_sentence, state_match, simplify = TRUE)
No_of_cases <- str_extract_all(news_sentence, "\\d+", simplify = TRUE)
x <- length(States)
y <- length(No_of_cases)
No_of_cases <- No_of_cases[1,(y-x+1):y]
covid_cases <- data.frame(New_states= t(States), No_of_cases)
covid_cases$No_of_cases <- as.numeric(covid_cases$No_of_cases)

states_df$states_affected <- gsub("FCT","Federal Capital Territory", states_df$states_affected)

```

## Project Objective

* NCDC Nigeria provides daily [Covid updates](https://covid19.ncdc.gov.ng/report/ "NCDC Nigeria Covid Daily Report").
* Online tables are easily downloaded with MS Excel.  
* However, this presentation also shares the current tables and the latest "Highlights".  





## Nigeria Covid-19 Status 


* Largest number of cases are in country's main hubs.

* States with the lowest number of cases also have low testing (not causation).


```{r Cases Overview,  message=FALSE, warning=FALSE, echo = FALSE}

ss <- states_df$states_affected
nn <- states_df$lab_confirmed_cases
bb <- c(0, 1000, 5000, 10000, 50000, 100000)
cc <- c("< 1k","1k to 5k", "5k to 10k", "10k to 50k", "> 50k")
dd <- paste("As reported by NCDC Nigeria on", pub_date)

map_ng(region = ss, x = nn, breaks = bb, col = 'red', categories = cc, 
       show.text = FALSE, title = "Distribution of Covid-19 Cases in Nigeria", 
       caption = dd)

```


## Cases by States 

* Lagos State has the highest number confirmed cases followed by the capital. 


```{r Total Cases, message=FALSE, warning=FALSE, echo = FALSE}

ggplot(states_df, aes(x = lab_confirmed_cases, y = reorder(states_affected, lab_confirmed_cases))) +
  geom_col()+
  theme(panel.background = element_rect(fill="white"),
        axis.ticks = element_blank(),
        axis.text = element_text(size=8),
        axis.title = element_text(size=12, face="bold"))+
  xlab("Number of lab. confirmed covid cases") + ylab("States")+
  labs(title = "Total Number of Lab. Confirmed Covid Cases by States",
       subtitle = "Official number of Covid-19 cases from inception",
       caption = dd)
```

## Discharged vs. Admissions vs. Deaths Cases. 

* Most of the officially confirmed cases were discharged. 

```{r Official Deaths vs. Admissions vs. Discharge, message=FALSE, warning=FALSE, echo = FALSE}


# Use this to show number of deaths vs. admissions vs discharge
df <- states_df[,3:5] %>% summarize_all(sum) %>%
  pivot_longer(cols = names(.))
df$name <- str_to_title(df$name)

plot_ly(type='pie', labels=df$name, 
        values= df$value, textinfo='label+percent',
        showlegend = FALSE, 
        textfont = list(size = 16),
        rotation = 90,
        title = 'Discharged vs. Admissions vs. Deaths',
        titlefont = list(size =25))
```

## Last Reported Number of Cases

* Given below are the lastest official results released by the NCDC Nigeria. 

```{r Last Reported Nummber of Cases, message=FALSE, warning=FALSE, echo = FALSE}

# Show most recent number of cases
ee <- paste("Confirmed cases on the", pub_date)

# Show total number cases in the country
ggplot(covid_cases, aes(x = No_of_cases, y = reorder(New_states, No_of_cases))) +
  geom_col()+
  theme(panel.background = element_rect(fill="white"),
        axis.ticks = element_blank(),
        axis.text = element_text(size=10),
        axis.title = element_text(size=12, face="bold"))+
  xlab("Number of lab. confirmed covid cases") + ylab("States")+
  labs(title = ee,
       subtitle = "Last official reported cases",
       caption = dd)

```


## The End
